---
title: "Tutorial: QHScrnomo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE, 
  message = FALSE,
  comment = "#>"
)
```

## Introduction

The `QHScrnomo` provides functions for fitting and predicting a competing risk model, creating a nomogram, k-fold cross validation, calculating the discrimination metric, and drawing calibration curve. This vignette will walk a reader through the various functions available.

## Setup

Before going through the tutorial, load {QHScrnomo}.

```{r message = FALSE, warning=FALSE, results='hide'}
library(QHScrnomo)
sessionInfo()
```

## Example data set

We'll be using the `prostate.dat` data set throughout this example.

* This is an artificial prostate cancer dataset used for illustrating the usages of functions in `QHScrnomo`

* This is a data frame with 2000 observations on the following 9 variables:

    - `UNIQID`: patient id
    - `TX`: Treatment options of prostate cancer with levels `EBRT`, `PI`, `RP`
    - `PSA`: Pre-treatment PSA levels
    - `BX_GLSN_CAT`: Biopsy Gleason Score Sum. a factor with levels 1 for 2-6 2 for 7 and 3 for 8-10
    - `CLIN_STG`: Clinical stage with levels `T1`, `T2`, `T3`
    - `Age`: Age at treatment date
    - `RACE_AA`: patient ethnicity.a factor with levels 0 for other and 1 for African American
    - `TIME_EVENT`: follow up time in months
    - `EVENT_DOD`: follow status, `0` - censored, `1` - died of prostate cancer, `2` - died of other causes
    
```{r}
data("prostate.dat")
str(prostate.dat)
```

## Fit a competing risk model

The function `crr.fit` uses a Cox proportional hazards regression model constructed from `cph` in `rms` library (by Frank Harrell).

```{r}
dd <- datadist(prostate.dat)
options(datadist = "dd")
prostate.f <- cph(Surv(TIME_EVENT,EVENT_DOD == 1) ~ TX  + rcs(PSA,3) +
           BX_GLSN_CAT + CLIN_STG + rcs(AGE,3) +
           RACE_AA, data = prostate.dat,
           x = TRUE, y= TRUE, surv=TRUE, time.inc = 144)
prostate.crr <- crr.fit(prostate.f, cencode = 0, failcode = 1)
summary(prostate.crr)
```

## Create nomogram

```{r fig.width=7, fig.height=5}
prostate.g <- Newlabels(prostate.crr,
                        c(TX = 'Treatment options', 
                          BX_GLSN_CAT = 'Biopsy Gleason Score Sum',
                          CLIN_STG = 'Clinical stage'))
nomogram.crr(prostate.g,
             failtime = 120,
             lp=FALSE,
             xfrac=0.65,
             fun.at = seq(0.2, 0.45, 0.05),
             funlabel = "Predicted 10-year cumulative incidence")
```

```{r}
# output a math formula
sas.cmprsk(prostate.crr, time = 120)
```


## Model evaluation

```{r}
# anova table
anova(prostate.crr)
```

```{r}
# prediction from 10-fold cross validation
prostate.dat$preds.tenf <- tenf.crr(prostate.crr, time=120, fold = 10)
str(prostate.dat$preds.tenf)
```

```{r}
## calculate the CRR version of concordance index
with(prostate.dat, cindex(preds.tenf,
                          ftime = TIME_EVENT,
                          fstatus =EVENT_DOD, type = "crr"))["cindex"]
```

```{r fig.width=5, fig.height=5}
## generate the calibration curve for predicted 10-year cancer
## specific mortality
with(prostate.dat,
     groupci(
         preds.tenf, 
         ftime = TIME_EVENT,
         fstatus =EVENT_DOD, g = 5, u = 120,
         xlab = "Nomogram predicted 10-year cancerspecific mortality",
         ylab = "Observed predicted 10-year cancerspecific mortality")
)
```

